From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-23 paper.Rmd
WIP

---
title: "From noise to knowledge: how randomness generates novel phenomena and information"
author: "Carl Boettiger"
---

```{r include=FALSE, message=FALSE}
knitr::opts_chunk$set(include = FALSE)
library(hrbrthemes)
library(tidyverse)
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
```

@@

In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.

@@

@@

@@

```{r}
read_csv("appendix/quasicycles.csv") %>%
  gather(species, pop, -t, -sigma) %>%
  ggplot(aes(t, pop, col=species)) + geom_line() + 
  facet_wrap(~sigma) + scale_color_ptol() + 
  labs(x="Time", 
       y="Population density",
       title="Quasi-cycles created by stochastic resonance",
       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
```

- Disease models

@@

@@

p1 <- tipping %>%
  select(-x) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y") + theme_ipsum() + 
  labs(title="A: Early warning signs")

p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
  scale_y_log10() + 
  theme_ipsum()+ labs(title="B: Population density",
                      caption = "Vertical red dashed line indicates tipping point location")

>>>>

---
title: "From noise to knowledge: how randomness generates novel phenomena and information"
author:
  - name: "Carl Boettiger"
    affiliation: a, 1
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"

abstract: |
  In this paper, I provide a brief synthesis of ecological research in stochastic phenomena. Noise, as the term itself suggests, is most often seen a nuisance to ecological insight, a inconvenient reality that must be acknowledged, a haystack that must be stripped away to reveal the processes of interest underneath. Yet despite this well-earned reputation, noise is often interesting in its own right: noise can induce novel phenomena that could not be understood from some underlying determinstic model alone.  Nor is all noise the same, and close examination of differences in in frequency, color or magnitude can reveal insights that would otherwise be inaccessible.  While an exhaustive review is beyond the scope of this piece, I endeavor to highlight some of the most interesting examples of both noise-induced phenomena and noise-derived information, and point the way towards yet uncharted territory.

keywords:
  - Stochasticity
  - Demographic noise
  - Environmental noise
  - Colored noise
  - quasi-cycles

lineno: true

bibliography: refs.bib
output: rticles::elsevier_article
---

```{r include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(hrbrthemes)
library(ggthemes)
library(tidyverse)
#ggplot2::theme_set(hrbrthemes::theme_ipsum())
```

@@

@@

![Figure 1: demographic stochasticity arises from discrete transitions  between states in in a Markov process.  Here, a simple birth-death process has transition rates determined by the current population size, _n_.](markov.png)

@@

<!-- Identify as beyond the scope of this paper -->

@@

```{r}
read_csv("appendix/quasicycles.csv") %>%
  gather(species, pop, -t, -sigma) %>%
  ggplot(aes(t, pop, col=species)) + geom_line() + 
  facet_wrap(~sigma) + scale_color_ptol() + 
  labs(x="Time", 
       y="Population density",
       title="Stochasticicity induces quasi-cycles",
       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
```

@@

**Persistence and coexistence**

One of the most fundamental aspects of stochasticty are the implications for long-term persistence of any given population and the stable coexistence of interacting species.

@@

p1 <- tipping %>%
  select(-x) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y") +
  #theme_ipsum() +
  labs(title="A: Early warning signs")

p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
  scale_y_log10() + 
  #theme_ipsum() + 
  labs(title="B: Population density",
       caption = "Vertical red dashed line indicates tipping point location")
